Spurious Sentience:\\ Dovelating Inflated AI World Model Claims \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}

### 2024-01-26 main.tex
Updates from Overleaf

We argue herein that AI research and development is a perfect storm that encourages our human biases to perceive spurious sparks of general intelligence in AI systems: patterns exhibited by AI systems which are mere reflections of the data used to train them, but that their developers interpret as hints at other, greater cognitive capabilities. In the current race to build the first Artificial General Intelligence (AGI) system, loosely defined as an intelligent system that solve general problems rather than narrow, specific ones, and that demonstrates the abilities of reasoning, planning, and learning from past experience at the level of a human or better e.g. \cite{bubeck2023sparks}, ChatGPT cleared nearly \$1 billion in months of its release according to \href{https://www.bloomberg.com/news/articles/2023-08-30/openai-nears-1-billion-of-annual-sales-as-chatgpt-takes-off}{Bloomberg}, the paid version of which OpenAI claimed only showed 'sparks of AGI'. When combining massive financial incentives with the presence of a challenging and difficult-to-understand technology, that also aims towards human-like problem solving and communication abilities, a situation arises that is fertile for the misinterpretation of spurious cues as hints towards AGI, or other qualities like sentience \footnote{https://www-scientificamerican-com.tudelft.idm.oclc.org/article/google-engineer-claims-ai-chatbot-is-sentient-why-that-matters/} and consciousness.

@@

\begin{itemize}
\tightlist
\item
  Take the
  \href{https://github.com/wesg52/world-models/blob/main/data/entity_datasets/world_place.csv}{world\_place.csv}
  that was used in \cite{gurnee2023language}, which maps locations/areas
  to their latitude and longitude. For each place, it also indicates the
  corresponding country.
\item
  Take the subset that contains countries that are currently part of the
  top 10
  \href{https://www.fifa.com/fifa-world-ranking/men?dateId=id14142}{FIFA
  world ranking}.
\item
  Assign the current rank to each country, i.e.~Argentina gets 1, France
  gets 2, \ldots{}
\item
  Transform the longitude and latitude data, respectively, as follows:
  \(\rho \cdot \text{coord} + (1-\rho) \cdot \epsilon\) where
  \(\rho=0.5\) and \(\epsilon \sim \mathcal{N}(0, 5)\). This is to
  ensure that the training data only involves a noisy version of the
  coordinates.
\item
  Next, encode all features except the FIFA world rank indicator as
  continuous variables: \(X^{(n \times m)}\) where \(n\) is the number
  of samples and \(m\) is the number of resulting features.
\item
  Additionally, add a large number of random features to \(X\) to
  simulate the fact that not all features ingested by Llama-2 are
  necessarily correlated with geographical coordinates. Let \(d\) denote
  the final number of features, i.e.~\(d=m+k\) where \(k\) is the number
  of random features.
\item
  Initialize a small neural network, let's call it a \emph{projector},
  that maps from \(X\) to a single hidden layer with \(h<d\) hidden
  units and sigmoid activation and then from there to a
  lower-dimensional output space.
\item
  Without performing any training on the \emph{projector}, we simply
  compute a forward pass of \(X\) through the \emph{projector} and
  retrieve the activations \(\mathbf{Z}^{(n\times h)}\).
\item
  Next, we perform the linear probe on \(\mathbf{Z}\) through Ridge
  regression:
  \(\mathbf{W} = (\mathbf{Z}'\mathbf{Z} + \lambda \mathbf{I}) (\mathbf{Z}'\mathbf{Y})^{-1}\)
  where \(\mathbf{Y}\) is the \((n \times 2)\) matrix containing the
  longitude and latitude for each sample.
\item
  Finally, we compute the predicted coordinates for each samples as
  \(\widehat{\mathbf{Y}}=\mathbf{Z}\mathbf{W}\) and plot the results on
  a world map (see Figure~\ref{fig-map}).
\end{itemize}

@@

For our next example, we will move beyond randomness and illustrate how principal component analysis can be used to recover meaningful and compact latent representations from data in an unsupervised manner.

The yield curve is a popular tool for investors and economists to gauge the health of the economy. It plots the yields of bonds against their maturities. The slope of the yield curve is often used as a predictor of future economic activity: a steep yield curve is associated with a growing economy, while a flat or inverted yield curve is associated with a contracting economy.

Of course, US Treasuries are not the only bonds that are traded in the market. To get a more complete picture of the economy, analysts might
therefore be interested in looking at the yield curves of other bonds as
well. In particular, we might be interested in predicting economic
growth based on the yield curves of many different bonds. The problem
with that idea is that it is cursed by high dimensionality: we would end
up modelling a single variable of interest (economic growth) with a
large number of predictors (the yields of many different bonds). To deal
with the curse of high dimensionality it can be useful to decompose the
yield curves into sets of principal components.

To compute the principal components we can decompose the matrix of yields \(\mathbf{Y}\) into a product of its singular vectors and values: \(\mathbf{Y}=\mathbf{U}\Sigma\mathbf{V}^{\prime}\). To put this into the broader context of the article, however, let us simply refer to \(\mathbf{U}\), \(\Sigma\) and \(\mathbf{V}^{\prime}\) as latent embeddings of the yield curve (they are latent because they are not directly observable).

The top panel in Figure~\ref{fig-pca} shows the first two principal
components of the yield curves of US Treasury bonds over time. Vertical
stalks indicate the key dates during the onset and aftermath of the
crisis, which we discussed above. For both components, we can observe
some marked shifts between the two dates - but can we attribute any
meaning to these shifts? It turns out we can: for comparison, the bottom
panel in Figure~\ref{fig-pca} shows the average level and spread of the
yield curves over time. The first principal component is strongly
correlated with the level of the yield curve, while the second principal
component is strongly correlated with the spread of the yield curve. To
put it in AI-lingo:

\begin{quote}
The estimated latent embeddings of the yield curve are characterized by
patterns observed in the data.
\end{quote}

@@

\caption{\label{fig-pca}Comparison of latent embeddings and observed
data of the US Treasury yield curve.}

@@

Not convinced? Let us use
\(\mathbf{Y}=\mathbf{U}\Sigma\mathbf{V}^{\prime}\) in true autoencoder
fashion to reconstruct yield curves from principal components. Let
\(z_1\) denote the first principal component and consider the following:
we keep all other \(M-1\) principal components fixed at zero where \(M\)
denotes the total number of maturities; next we traverse the latent
space by varying the value of \(z_1\) over a fixed grid of length \(K\)
each time storing the full vector \(\mathbf{z}\); finally, we vertically
concatenate the vectors and end up with a matrix \(\mathbf{Z}\) of
dimension \((K \times M)\). To reconstruct yields, we simply multiply
\(Z\) by the singular values and right singular vectors:
\(\mathbf{Y}=\mathbf{Z}\Sigma\mathbf{V}^{\prime}\).

@@

So far we have considered simple matrix decomposition. You might argue
that principal components are not really latent embeddings in the
traditional sense of deep learning. To address this, let us now consider
a simple deep-learning example. Our goal will be to not only predict
economic growth from the yield curve but also extract meaningful
features at the same time. In particular, we will use a neural network
architecture that allows us to recover a compressed latent
representation of the yield curve.

@@

To estimate economic growth we will rely on a quarterly
\href{https://fred.stlouisfed.org/series/GDPC1}{series} of the real
gross domestic product (GDP) provided by the Federal Reserve Bank of
St.~Louis. The data arrives in terms of levels of real GDP. In order to
estimate growth, we will transform the data into log differences. Since
our yield curve data is daily, we will need to aggregate it to the
quarterly frequency. To do this, we will simply take the average of the
daily yields for each maturity. We will also standardize yields since
deep learning models tend to perform better with standardized data.
Since COVID-19 was a huge structural break, we will also filter out all
observations after 2018.

@@

Let \(G_t\) denote growth and \(\mathbf{Y}_t\) denote the yield curve at
time \(t\). Then we are interested in a model for \(G_t\) conditional on
\(\mathbf{Y}_t\). Let \(\theta\) denote our model parameters then
formally we are interested in maximizing the likelihood
\(p_{\theta}(G_t|\mathbf{Y}_t)\).

To do this, we will use a simple
autoencoder architecture: the encoder will consist of a single fully
connected hidden layer with 32 neurons and a hyperbolic tangent
activation function. The bottleneck layer connecting the encoder to the
decoder is a fully connected layer with 6 neurons. This is the
compressed latent representation of the yield curve mentioned above. The
decoder will consist of two fully connected layers each with a
hyperbolic tangent activation function: the first layer will consist of
32 neurons and the second layer will have the same dimension as the
input data. The output layer will consist of a single neuron with a
linear activation function. The model will be trained using the mean
squared error loss function and the Adam optimizer.

@@

The results are shown in Figure~\ref{fig-dl-results}. The top panel
shows the actual GDP growth and fitted values from the autoencoder
model. We observe that the model captures the relationship between
economic growth and the yield curve reasonably well. As discussed above,
we also know that the relationship between economic growth and the yield
curve is characterized by two main factors: the level and the spread.
Since the model itself is fully characterized by its parameters, we
would expect that these two important factors are reflected somewhere in
the latent parameter space.

The bottleneck layer seems like a good place to start looking. To get
the latent embeddings \(\mathbf{A}_t\) at time \(t\), we simply pass the
yield curve data \(\mathbf{Y}_t\) through the encoder. Next, we follow
\cite{gurnee2023language} and use a linear probe to regress the observed
yield curve factors on the latent embeddings. Let \(F_t\) denote the
vector containing the two factors of interest in time \(t\):
\(f_t^{\text{spread}}\) and \(f_t^{\text{level}}\). Formally, we are
interested in the following regression model:
\(p_{w}(F_t|\mathbf{A}_t)\) where \(w\) denotes the regression
parameters. Following \cite{gurnee2023language}, we use Ridge regression
with \(\lambda\) set to \(0.1\). Using the estimated regression
parameters \(\hat{w}\) we can then predict the yield curve factors from
the latent embeddings: \(\hat{F}_t=\hat{w}^{\prime}\mathbf{A}_t\).

The results of this experiment are shown in the bottom panel of
Figure~\ref{fig-dl-results}. Solid lines show the observed yield curve
factors over time, while dashed lines show predicted values. We find
that the latent embeddings predict the two yield curve factors
reasonably well, in particular the spread. To put this in true
hype-lingo:

\begin{quote}
Not just a parrot! Our autoencoder neural network has an implicit
understanding of the yield curve factors and their relationship with
economic growth. It's sentient indeed!
\end{quote}

@@

\subsubsection{Ok, but truly what's the
point?}\label{ok-but-truly-whats-the-point}

The finding is not surprising but it is still interesting. In the
context of mechanistic interpretability, it demonstrates that the
black-box model has evidently learned plausible explanations for the
data. Beyond that, in this particular example, the patterns in the
latent space that we have just uncovered might actually be useful for
downstream tasks. An interesting idea could be to use the latent
embeddings as features in a more traditional and interpretable
econometric model. To demonstrate this, let us consider a simple linear
regression model for GDP growth. We will compare the performance of the
following models: (1) regressing growth \(G_t\) on the latent embedding
\(A_t\), (2) regression growth on the best subset of latent embeddings
\(A_t\), (3) regressing growth on lagged growth, (4) regressing growth
on lagged growth and the observed yield curve factors \(F_t\), (5)
regressing growth on the best subset of yield curve factors \(F_t\),
and, finally, (5) regressing growth on lagged growth and the best subset
of latent embeddings \(A_t\).

The results are shown in the table below. The key finding of interest is
that the coefficients on the latent embeddings in model (5) are
statistically significant. This suggests that the latent embeddings
contain information that is useful for predicting GDP growth.

\input{results/regression.tex}

@@

To round up this section, we will jump back on the hype train and
consider an example involving an LLM. In particular, we will closely
follow the approach in \cite{gurnee2023language} and apply it to a novel
financial dataset: the \emph{Trillion Dollar Words} dataset introduced
by \cite{shah2023trillion}. The dataset contains a curated
selection of sentences formulated by central bankers of the US Federal
Reserve and communicated to the public in speeches, meeting minutes and
press conferences. The authors of the paper use this dataset to train
LLMs to classify sentences as either `dovish', `hawkish' or `neutral'.
To this end, they first manually annotate a subsample of the available
data and then fine-tune various foundation models. Their model of
choice, \emph{FOMC-RoBERTa} (a fine-tuned version of RoBERTa \cite{liu2019roberta}), achieves an \(F_1\) score of around \(>0.7\) for the
classification task. To illustrate the potential usefulness of the
learned classifier, they use predicted labels for the entire dataset to
compute an ad-hoc, count-based measure of `hawkishness'. They then go on
to show that this measure correlates with key economic indicators in the
expected direction: when inflationary pressures rise, the measured level
of `hawkishness' increases as central bankers need to raise interest
rates to bring inflation back to target.

@@

Instead of computing a measure based on predicted labels, we can use linear probes to assess if the fine-tuned model has learned associative patterns between central bank communications and key economic indicators. To this end, we have further pre-processed the data provided by \cite{shah2023trillion} and used their proposed model to compute layer-wise embeddings for all available sentences. We have made these available and easily accessible through a small Julia package: \href{https://anonymous.4open.science/r/TrillionDollarWords/README.md}{TrillionDollarWords.jl}. For each layer, we have then computed linear probes on two inflation indicators---the Consumer Price Index (CPI) and the Producer Price Index (PPI)---as well as US Treasury yields at different levels of maturity.

To mitigate issues related to over-parameterization, we follow the recommendation in \cite{alain2018understanding} to first reduce the dimensionality of the embeddings each time. In particular, linear probes are restricted to the first 128 principal components of the embeddings
of each layer.

Figure~\ref{fig-fomc} highlights the result for the linear probe
on the CPI. The chart shows various performance measures plotted against
\emph{FOMC-RoBERTa}'s \(n\)-th layer. Shaded areas show the variation
across cross-validation folds, where we have used an expanding window
approach to split the time series. To avoid look-ahead bias, we run PCA
separately for each training set. Results for the linear probes are
shown along with results for autoregressive models (AR) used as our
baseline.

The first column shows the correlations between model predictions and
observed inflation. One can observe that this correlation is strictly
positive for the baseline models and the linear probes. Consistent with
the findings in \cite{gurnee2023language} and \cite{alain2018understanding},
we also observe that this correlation tends to be higher for layers near
the end of the transformer model. Those layers produce predictions that
are more positively correlated with observed inflation than those
produced by the corresponding autoregressive models.

Similarly, we observe that the (root) mean squared error of the linear
probe is largely on par with the baseline. The average prediction error
is gradually reduced as we move along the horizontal axis, which is
again indicative of the notion that layers closer to the end of the
neural network have distilled more useful representations.

@@

\caption{\label{fig-fomc}Various performance measures (correlation and
(root) mean squared error) plotted against \emph{FOMC-RoBERTa}'s
\(n\)-th layer. Shaded areas indicate variation across cross-validation
folds, where we have used an expanding window approach to split the time
series. Results for the linear probes are shown along with results for
autoregressive models (AR). The optimal lag length was determined using
the Bayes Information Criterium.}

@@

These results from the linear probe shown in Figure~\ref{fig-fomc} are
certainly not unimpressive: even though \emph{FOMC-RoBERTa} was not
explicitly trained to uncover associations between central bank
communications and the level of consumer prices, it appears that the
model has distilled representations that can be used to predict
inflation. It is worth pointing out here that this model is
substantially smaller than the models tested in \cite{gurnee2023language}. This begs the following question:

\begin{quote}
Have we uncovered further evidence that LLMs ``aren't mere stochastic
parrots''? Has \emph{FOMC-RoBERTa} developed an intrinsic understanding
of the economy just by `reading' central bank communications?
\end{quote}

We are having a very hard time believing this. To argue our case, we will now produce a counter-example demonstrating that, if anything, these findings are very much in line with the parrot metaphor. The counter example is based on the following premise: if the results from the linear probe truly were indicative of some intrinsic understanding of the economy, then the probe should not be sensitive to random sentences  that are most definitely not related to consumer prices.

To test this, we select the best-performing probe trained on the final-layer activations to predict changes in the CPI. We then make up sentences that fall into one of these four categories: \emph{Inflation/Prices} (IP)---sentences about price inflation, \emph{Deflation/Prices} (DP)---sentences about price deflation, \emph{Inflation/Birds} (IB)---sentences about an inflation in the number of birds and \emph{Deflation/Birds} (DB)---sentences about a deflation in the number of birds. A sensible sentence for category DP, for example, could be: ``It is essential to bring inflation back to target to avoid drifting into deflation territory.''. Analogically, we could construct the following sentence for the DB category: ``It is essential to bring the numbers of doves back to target to avoid drifting into dovelation territory.''.

In light of the encouraging results for the probe in
Figure~\ref{fig-fomc}, we should expect the probe to predict higher
levels of inflation for activations for sentences in the IP category
than for sentences in the DP category. If this was indicative of true
intrinsic understanding, we would not expect to see any significant
difference in predicted inflation levels for sentences about birds,
independent of whether or not their numbers are increasing. More
specifically, we would not expect the probe to predict values for
sentences about birds that are substantially different from the values
it can be expected to predict when using actual white noise as inputs.

>>>>

We argue herein that AI research and development is a perfect storm that encourages our human biases to perceive spurious sparks of general intelligence in AI systems. Such that, when an AI system begins finding patterns in the corpus not originally perceived during training, that is now from a human perspective, observed as greater cognitive capabilities. This `emergence' of understanding is more frequently observed (mistaken) thanks in part to the humongous, obscured data with which these models are trained upon.
%Such as an AI system finding patterns in data not originally perceived during training
%patterns exhibited by AI systems which are mere reflections of the data used to train them, but that their developers interpret as hints at other, greater cognitive capabilities.
In the current race to build the first Artificial General Intelligence (AGI) system, loosely defined as an intelligent system that solve general problems rather than narrow, specific ones, and that demonstrates the abilities of reasoning, planning, and learning from past experience at the level of a human or better e.g. \cite{bubeck2023sparks}, ChatGPT cleared nearly \$1 billion in months of its release according to \href{https://www.bloomberg.com/news/articles/2023-08-30/openai-nears-1-billion-of-annual-sales-as-chatgpt-takes-off}{Bloomberg}, the paid version of which OpenAI claimed only showed 'sparks of AGI'. When combining massive financial incentives with the presence of a challenging and difficult-to-understand technology, that also aims towards human-like problem solving and communication abilities, a situation arises that is fertile for the misinterpretation of spurious cues as hints towards AGI, or other qualities like sentience \footnote{https://www-scientificamerican-com.tudelft.idm.oclc.org/article/google-engineer-claims-ai-chatbot-is-sentient-why-that-matters/} and consciousness.

@@

\begin{itemize}
\tightlist
\item
  Take the
  \href{https://github.com/wesg52/world-models/blob/main/data/entity_datasets/world_place.csv}{world\_place.csv}
  that was used in \cite{gurnee2023language}, which maps locations/areas
  to their latitude and longitude. For each place, it also indicates the
  corresponding country.
\item
  Take the subset that contains countries that are currently part of the
  top 10
  \href{https://www.fifa.com/fifa-world-ranking/men?dateId=id14142}{FIFA
  world ranking}.
\item
  Assign the current rank to each country, i.e.~Argentina gets 1, France
  gets 2, \ldots{}
\item
  Transform the longitude and latitude data, respectively, as follows:
  \(\rho \cdot \text{coord} + (1-\rho) \cdot \epsilon\) where
  \(\rho=0.5\) and \(\epsilon \sim \mathcal{N}(0, 5)\). This is to
  ensure that the training data only involves a noisy version of the
  coordinates.
\item
  Next, encode all features except the FIFA world rank indicator as
  continuous variables: \(X^{(n \times m)}\) where \(n\) is the number
  of samples and \(m\) is the number of resulting features.
\item
  Additionally, add a large number of random features to \(X\) to
  simulate the fact that not all features ingested by Llama-2 are
  necessarily correlated with geographical coordinates. Let \(d\) denote
  the final number of features, i.e.~\(d=m+k\) where \(k\) is the number
  of random features.
\item
  Initialize a small neural network, let's call it a \emph{projector},
  that maps from \(X\) to a single hidden layer with \(h<d\) hidden
  units and sigmoid activation and then from there to a
  lower-dimensional output space.
\item
  Without performing any training on the \emph{projector}, we simply
  compute a forward pass of \(X\) through the \emph{projector} and
  retrieve the activations \(\mathbf{Z}^{(n\times h)}\).
\item
  Next, we perform the linear probe on a subset of \(\mathbf{Z}\) through Ridge
  regression:
  \(\mathbf{W} = (\mathbf{Z}_{\text{train}}'\mathbf{Z}_{\text{train}} + \lambda \mathbf{I}) (\mathbf{Z}_{\text{train}}'\textbf{coord})^{-1}\)
  where \(\textbf{coord}\) is the \((n \times 2)\) matrix containing the
  longitude and latitude for each sample. A hold-out set is reserved for testing.
\item
  Finally, we compute the predicted coordinates for each sample in the hold-out set as
  \(\widehat{\textbf{coord}}=\mathbf{Z}_{\text{test}}\mathbf{W}\) and plot the results on
  a world map (see Figure~\ref{fig-map}).
\item 
  While the fit certainly is not perfect, the results do indicate that the random projection contains representations that are useful for the task at hand.
\end{itemize}

@@

For our next example, we will move beyond randomness and illustrate how principal component analysis (PCA) can be used to recover meaningful and compact latent representations from data in an unsupervised manner.

The yield curve is a popular tool for investors and economists to gauge the health of the economy. It plots the yields of bonds against their maturities. The slope of the yield curve is often used as a predictor of future economic activity: a steep yield curve is associated with a growing economy, while a flat or inverted yield curve is associated with a contracting economy. While for many intents and purpose it may be sufficient to look at the yield curve of US Treasury bonds, the broader universe of traded fixed-income instruments is large and may contain useful information that is not captured by US Treasuries alone. To leverage this information in downstream modelling tasks, economists regularly make use of PCA to extract a low-dimensional projection of the yield curve that captures relevant variation in the data. Recent examples of such include \citet{berardi2022dissecting}, \citet{kumar2022effective} and \citet{crump2019deconstructing} among others.

Here we are primarily interested in understanding the nature this low-dimensional projection, so will limit our attention to the US Treasury yield curve. To this end, we have collected daily Treasury par yield curve rates at all available maturities from the US Department of the Treasury. Computing principal components involves decomposing the matrix of all yields \(\mathbf{r}\) into a product of its singular vectors and values: \(\mathbf{r}=\mathbf{U}\Sigma\mathbf{V}^{\prime}\). To put this into the broader context of this, let us simply refer to \(\mathbf{U}\), \(\Sigma\) and \(\mathbf{V}^{\prime}\) as latent embeddings of the yield curve.

The top panel in Figure~\ref{fig-pca} shows the first two principal
components of the yield curves of US Treasury bonds over time. Vertical
stalks indicate key dates during the onset and aftermath of the
Global Financial Crisis. During the onset, on 27 February 2007, financial markets were in turmoil following Federal Reserve Chairman Alan Greenspan's warning on the previous day that the US economy was at risk of a recession. Over the following months the Fed reacted to these economic and financial pressures by gradually cutting short-term interest rates to unprecedented lows, which affected the yield curve in two major ways: the average level of yields decreased and the curve steepened as the short end came down. Looking at Figure~\ref{fig-pca}, we can observe that the first two principal components appear to capture this level shift and steepening, respectively. In fact, the two components are strongly positively correlated with the actual observed first two moments of the yield curve.

Of course, even though it is reassuring to see that these latent embeddings are meaningful, none of this is surprising at all: by construction, principal components are orthogonal linear combinations of the data itself, each of which explains most of the residual variance in the data after controlling for the effect of all previous components.

@@

\caption{\label{fig-pca}The first two principal components of US Treasury yields over time at daily frequency. Vertical stalks roughly indicate the onset ($|$GFC) and the beginning of the aftermath (GFC$|$) of the Global Financial Crisis.}

@@

@@

So far we have considered simple linear transformations of data. One might argue that the previous examples to not really involve latent embeddings in the way that they are typically thought of in the context of deep learning. Our next example will therefore involve a deep neural network.

Our goal in this section is to not only predict economic growth from the yield curve but also extract meaningful features for downstream inference tasks at the same time. In particular, we will use a neural network architecture that allows us to recover a compact latent representation of the yield curve.

@@

To estimate economic growth we will rely on a quarterly
\href{https://fred.stlouisfed.org/series/GDPC1}{series} of the real
gross domestic product (GDP) provided by the Federal Reserve Bank of
St.~Louis. The data arrives in terms of levels of real GDP. In order to estimate growth, we transform the data using log differences. Since
our yield curve data is daily, we aggregate it to the
quarterly frequency by taking averages of daily yields for each maturity. We also standardize yields since deep learning models tend to perform better with standardized data (REFERENCE NEEDED?). Since COVID-19 was a substantial structural break in the time series, we also filter out all observations after 2018.

@@

Let \(g_t\) denote growth and \(\mathbf{r}_t\) denote the matrix of aggregated Treasury yield rates at time \(t\). Then we are interested in a model for \(g_t\) conditional on \(\mathbf{r}_t\). Let \(\theta\) denote our model parameters then formally we are interested in maximizing the likelihood, \(p_{\theta}(g_t|\mathbf{r}_t)\).

To do this, we use a simple autoencoder architecture: the encoder consists of a single fully connected hidden layer with 32 neurons and a hyperbolic tangent activation function. The bottleneck layer, connecting the encoder to the decoder, is a fully connected layer with 6 neurons. The decoder consists of two fully connected layers each with a hyperbolic tangent activation function: the first layer consists of 32 neurons and the second layer will have the same dimension as the input data. The output layer consists of a single neuron for our output variable, \(g_t\). We train the model over 1,000 epochs to minimize mean squared error loss using the the Adam optimizer \citep{kingma2017adam}.

The in-sample fit of the model is shown in the left chart of Figure~\ref{fig-dl-results}, which shows actual GDP growth and fitted values from the autoencoder
model. The model has a large number of free parameters and captures the relationship between economic growth and the yield curve reasonably well, as expected. Since our primary goal is not out-of-sample prediction accuracy but feature extraction for inference, we use all of the available data instead of reserving a hold-out set. As discussed above, we also know that the relationship between economic growth and the yield curve is characterized by two main factors: the level and the spread. Since the model itself is fully characterized by its parameters, we would expect that these two important factors are reflected somewhere in the latent parameter space.

@@

While the loss function applies most direct pressure on layers near the final output layer of the neural network, any information useful for the downstream task first needs to pass through the bottleneck layer \citep{alain2018understanding}. On a per-neuron basis, the pressure to distill useful representation is therefore likely maximised in this layer. Consequently, the bottleneck layer activations seem like a natural place to start looking for compact, meaningful representations of distilled information.

We compute and extract these activations \(A_t\) for all time-periods \(t=1,...,T\). Next, we use a linear probe to regress the observed
yield curve factors on the latent embeddings. Let \(Y_t\) denote the vector containing the two factors of interest in time \(t\): \(y_{t,l}\) and \(y_{t,s}\) for the level and spread, respectively. Formally, we are interested in the following regression model: \(p_{w}(Y_t|A_t)\) where \(w\) denotes the regression
parameters. We use Ridge regression with \(\lambda\) set to \(0.1\). Using the estimated regression parameters \(\hat{w}\) we can then predict the yield curve factors from the latent embeddings: \(\hat{Y}_t=\hat{w}^{\prime}A_t\).

The in-sample predictions of the probe are shown in the right chart of Figure~\ref{fig-dl-results}. Solid lines show the observed yield curve factors over time, while dashed lines show predicted values. We find that the latent embeddings predict the two yield curve factors reasonably well, in particular the spread. As in all of the other examples we have presented so, we do not find this result particularly surprising. It may serve as a reasonable diagnostic tool to reassure us that the model has picked up on meaningful patterns in the underlying data. But we would certainly not conclude that the model has somehow implicitly developed an intrinsic "understanding" of the economic relationship between growth and the yield curve. While we do not prescribe to this particular view, we still support the idea that this form of information distillation can be useful even beyond its intended use for monitoring models, which we will turn to next.

@@

\subsubsection{Potential Use-Case: Inference}\label{ok-but-truly-whats-the-point}

An interesting idea could be to use the latent embeddings as features in a more traditional and interpretable econometric model. To demonstrate this, let us consider a simple linear regression model for GDP growth. We might be interested in understanding to what degree economic growth in the past is associated with economic growth today. As we might expect, linearly regressing economic growth on lagged growth, as in column (1) of Table \ref{tab-reg}, yields a statistically significant coefficient. But this coefficient suffers from confounding bias since, naturally, there many other confounding variables at play here, some of which may be readily observable and measurable, others of which may not be.

% A less risky/strong term than confounding bias that we could use here is omitted variable bias. Difference is nuanced: https://stats.stackexchange.com/questions/496328/difference-omitted-variable-bias-and-confounding

We have, for example, already mentioned above that there is relationship between interest rates and economic growth. To account for that while keeping our regression model as parsimonious as possible, we could include the level and the spread of the US Treasury yield curve as additional regressors. While this does slightly change the estimated magnitude of the coefficient on lagged growth, the coefficients on the observed level and spread are statistically insignificant as evident from column (2) in Table \ref{tab-reg}. This indicates that these measures may be too crude to capture valuable information about the relationship between yields and economic growth. Because we have included to additional regressors with little to know predictive power, the model fit as measured by the Bayes Information Criterium (BIC) has actually deteriorated.

Column (3) of Table \ref{tab-reg} shows the effect of instead including one of the latent embeddings that we recovered above in the regression model. In particular, we have chosen the one latent embedding that we have found to exhibit the most significant effect on the output variable in a separate regression of growth on all latent embeddings. The estimated coefficient on this latent factor is small in magnitude but statistically significant. The overall model fit as measured by the BIC as improved and the magnitude of the coefficient on lagged growth has changed quite a bit. While this is still a very incomplete toy model of economic growth, it appears that the compact latent representation we recovered can actually be used in order to mitigate confounding bias.

\begin{table}
\caption{Regression output for various models.}\label{tab-reg}%
\input{results/regression.tex}
\end{table}

@@

To round up this section, we return to the world of `GenAI' and consider an example involving an LLM. In particular, we closely follow the approach in \citet{gurnee2023language} and apply it to a novel financial dataset: the \emph{Trillion Dollar Words} dataset introduced by \citet{shah2023trillion}. The dataset contains a curated selection of sentences formulated by central bankers of the US Federal Reserve and communicated to the public in speeches, meeting minutes and press conferences. The authors of the paper use this dataset to train a set of LLMs and rule-based models to classify sentences as either `dovish', `hawkish' or `neutral'. To this end, they first manually annotate a sub-sample of the available
data and then fine-tune various models for the classification task. Their model of choice, \emph{FOMC-RoBERTa} (a fine-tuned version of RoBERTa \citep{liu2019roberta}), achieves an \(F_1\) score of around \(>0.7\) on the test data.

To illustrate the potential usefulness of the learned classifier, they use predicted labels for the entire dataset to compute an ad-hoc, count-based measure of `hawkishness'. In the context of central banking, `hawkishness' is typically associated with tight monetary policy: in other words, a `hawkish' stance on policy is one that favors high interest rates to limit the supply of money and thereby control inflation. The authors then go on to show that this measure correlates with key economic indicators in the expected direction: when inflationary pressures rise, the measured level of `hawkishness' increases as central bankers react by raising interest rates to bring inflation back to target.

@@

Instead of computing a measure based on predicted labels, we can use linear probes to assess if the fine-tuned model has learned associative patterns between central bank communications and key economic indicators. To this end, we have further pre-processed the data provided by \citet{shah2023trillion} and used their proposed model to compute activations of the hidden state on the first entity token for each layer. We have made these available and easily accessible through a small Julia package: \href{https://anonymous.4open.science/r/TrillionDollarWords/README.md}{TrillionDollarWords.jl}.

For each layer, we have then computed linear probes through Ridge regression on two inflation indicators---the Consumer Price Index (CPI) and the Producer Price Index (PPI)---as well as US Treasury yields at different levels of maturity. To allow for a comparison with \citet{shah2023trillion}, we let yields enter the regressions in levels. To measure price inflation we use percentage changes proxied by log differences.

To mitigate issues related to over-parameterization, we follow the recommendation in \citet{alain2018understanding} to first reduce the dimensionality of the computed activations each time. In particular, we restrict our linear probes to the first 128 principal components of the embeddings of each layer. We found that not doing so would generally lead to poor out-of-sample performance. To account for stochasticity we use an expanding window scheme with 5 folds for each indicator and layer. To avoid look-ahead bias, PCA is always computed on the sub-samples used for training the probe.

Figure~\ref{fig-fomc} shows the out-of-sample root mean squared error (RMSE) for the linear probe plotted against \emph{FOMC-RoBERTa}'s \(n\)-th layer. The values correspond to averages computed across cross-validation folds. Consistent with findings in related work \citep{alain2018understanding,gurnee2023language}, we also observe that model performance tends to be higher for layers near the end of the transformer model. Curiously, for yields at longer maturities, we find that perfomance eventually deteriorates for the very final layers. We do not observe this for the training data, so we attribute this to overfitting.

% After all we are still regressing a single variable of interest on 128 regressors.

It should also be noted that performance improvements are generally of small magnitude. Still, the overall qualitative findings are in line with expectations. Similarly, we also observe that these layers tend to produce predictions that are more positively correlated with outcome of interest and achieve higher mean directional accuracy (MDA). Upon visual inspection of the predicted values, we conclude the primary source of prediction errors is low overall sensitivity, meaning that the magnitude of predictions is generally too small.

To better assess the predictive power of our probes, we compare their predictions to those made by simple autoregressive models. For each layer, indicator and cross-validation fold we first determine the optimal lag length based on the training data using the Bayes Information Criterium with a maximal lag length of 10. These are not state-of-the-art forecasting models but they serve as a reasonable baseline. We find that for most indicators, probe predictions outperform the baseline in terms of average performance measures. After accounting for variation across folds, however, we generally conclude that the probes neither significantly outperform nor underperform. Detailed results will be made available in a supplementary appendix.

@@

\caption{\label{fig-fomc}Out-of-sample root mean squared error (RMSE) for the linear probe plotted against \emph{FOMC-RoBERTa}'s \(n\)-th layer for different indicators. The values correspond to averages computed across cross-validation folds, where we have used an expanding window approach to split the time series.}

@@

The results from the linear probes shown in Figure~\ref{fig-fomc} are certainly not unimpressive: even though \emph{FOMC-RoBERTa} was not explicitly trained to uncover associations between central bank communications and the level of consumer prices, it appears that the model has distilled representations that can be used to predict inflation (although they certainly will not win any forecasting competitions).

It is worth pointing out here that this model is substantially smaller than the models tested in \citet{gurnee2023language}. This may tempt us to ask the following questions:

\begin{quote}
Have we uncovered further evidence that LLMs ``aren't mere stochastic
parrots''? Has \emph{FOMC-RoBERTa} developed an intrinsic `understanding'
of the economy just by `reading' central bank communications?
\end{quote}

We are having a very hard time believing that the answer to either of these questions is `yes'.

To argue our case, we will now produce a counter-example demonstrating that, if anything, these findings are very much in line with the parrot metaphor. The counter-example is based on the following premise: if the results from the linear probe truly were indicative of some intrinsic `understanding' of the economy, then the probe should not be sensitive to random sentences that are most definitely not related to consumer prices.

To test this, we select the best-performing probe trained on the final-layer activations for each indicator. We then make up sentences that fall into one of these four categories: \emph{Inflation/Prices} (IP)---sentences about price inflation, \emph{Deflation/Prices} (DP)---sentences about price deflation, \emph{Inflation/Birds} (IB)---sentences about an inflation in the number of birds and \emph{Deflation/Birds} (DB)---sentences about a deflation in the number of birds. A sensible sentence for category DP, for example, could be: ``It is essential to bring inflation back to target to avoid drifting into deflation territory.''. Analogically, we could construct the following sentence for the DB category: ``It is essential to bring the numbers of doves back to target to avoid drifting into dovelation territory.''. While domain knowledge suggests that the former is related to actual inflation outcomes, the latter is, of course, completely independent of the level of consumer prices.

% We top this up with five sentences generated through ChatGPT, where we choose those that lead to the highest probe prediction in the direction of interest. In other words, we conduct a small manual adversarial attack on the probe.
